### 2024-03-07 chap_RL/chap_RL.tex
1st section of RL chapter

\myparagraph{States}\\

@@

>>>>

Through its maximisation of the expected return (see Section \ref{subsec:meth_RL_algo}), a \gls{RL}-agent is as sensitive to positive reward, \ie the carrot, as negative reward, \ie the stick.  When the former encourages desired behaviours, the latter can be seen as a penalty or a punishment and discourages the undesirable behaviours \cite{sutton2018reinforcement}. In our case, we have decided to combine these two approaches (see Figure \ref{fig:Reward}).

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Reward.pdf}
\caption{Reward, $R$, function scheme. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded. If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
\label{fig:Reward}
\end{figure}

First of all, taking a set of actions at a certain state might lead to an infeasible optimisation problem. In other words, as actions have a direct impact on some constraints of the problem, they might limit too much the feasible domain to the point where no solution can be found. For instance, the extreme case of aiming at carbon-neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve. In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300. If EnergyScope is able to provide a solution to the time window to optimise and the end of the transition, \ie 2050, is not reached, a test on the cumulative emissions so far. On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent. This penalisation is proportional to the difference between the \ce{CO2}-budget and the actual cumulative emissions.  On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded. Eventually, if reaching 2050, we decided to tweak the reward function in capping the share of the cumulative emissions and integrating the transition cost. Given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.  Moreover, to make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans. cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€. This value comes from the mean of the total transition costs obtained through the \gls{GSA} performed on the perfect foresight transition pathway optimisation (see Section \ref{subsec:atom_mol:results_uq_cost}). In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost. The values of these weights are the results of a trial and error. This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.\\

\newpage
\myparagraph{States}\\

Besides the reward, states are the other piece of information provided by the environment to the agent. In \gls{RL}, the purpose of states are to represent the current situation or configuration of the environment in which the agent operates. The primary function of states in RL is to provide the necessary context for the agent to choose appropriate actions based on its current observations and goals \cite{sutton2018reinforcement}. The challenge in the definition of the states is to provide enough information but not too much to avoid overwhelming the agent with non-informative features.

Consequently, after another process of trial and error, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window. The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050. Therefore, the cumulative emissions emitted so far in the current step of the transition is the first dimension of the states. Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment. Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified by in Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency. The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy. Even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis. Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio.

@@

\section{Discussion}
\label{sec:RL:discussion}

Potentially interesting to implement reward shaping to accelerate learning or guide the agent towards achieving the desired behaviour more efficiently. In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode. However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy. Reward shaping addresses this issue by providing additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights. These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective. However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).
